Globally-centered autocovariances in MCMC

### 2020-09-03 Paper/main.tex
Reviewed first two sections. Heading to ze office.

\onehalfspacing
\begin{abstract}
Autocovariances are a fundamental quantity of interest in Markov chain Monte Carlo (MCMC) simulations with autcorrelation function (ACF) plots being often used as a visual tool to ascertain the performance of a Markov chain. Unfortunately, for slow mixing Markov chains, the empirical autocovariance can highly underestimate the truth. For multiple chain MCMC sampling,  we propose a globally-centered estimator of the autocovariance function (G-ACvF) that exhibits theoretical and empirical improvements. We show that the bias of the G-ACvF estimator is smaller than the bias of the current state-of-the-art.  The impact of this improved estimator is evident in three critical output analysis applications: (1) ACF plots, (2) estimates of the Monte Carlo asymptotic covariance matrix, and (3) estimates of the effective sample size. Under weak conditions, we establish strong consistency of an improved asymptotic covariance estimator, and obtain its large sample bias and variance.  
\end{abstract}

@@

Let $F$ be the target distribution with mean $\mu$ defined on a space $\X \subseteq \mathbb{R}^d$ equipped with a countably generated $\sigma$-field $\mathcal{B}(\X)$. For $s = 1, \dots, m$, let $\{X_{st}\}_{t\geq1}$ be the $s${th} Harris ergodic $F$-invariant Markov chain \citep[see][for definitions]{meyn:twee:2009} employed to learn characteristics about $F$. The process is covariance stationary so that
the ACvF at lag $k$ is defined as 
%
\[
    \Gamma(k) = \Cov_F(X_{s1}, X_{s\,1+k})= \mathbb{E}_F \left[(X_{s1} - \mu)(X_{s\,1+k} - \mu)^T \right]\,.
\]
Estimating $\Gamma(k)$ is critical to assessing the quality of the sampler and the reliability of Monte Carlo estimators. Let $\bar{X}_s = n^{-1} \sum_{t=1}^{n} X_{st}$ denote the Monte Carlo estimator of $\mu$ from the $s$th chain. The standard estimator for $\Gamma(k)$ is the sample autocovariance matrix at lag $k \geq 0$:
%
\begin{equation} \label{eq:empirical_ACvF}
    \hat{\Gamma}_s(k) = \dfrac{1}{n}\sum_{t=1}^{n-k} \left(X_{st} - \bar{X}_s \right) \left(X_{s\, t + k} - \bar{X}_s \right)^T\,,
\end{equation}
and for $k < 0$, $\hat{\Gamma}_s(k) = \hat{\Gamma}_s(-k)^T$. For a single-chain MCMC run, the estimator $\hat{\Gamma}_s(k)$ is used to construct ACF plots, to estimate the long-run variance of Monte Carlo estimators \citep{hannan:1970,dame:1991}, and to estimate effective sample size (ESS) \citep{kass:carlin:gelman:neal:1998,gong:fleg:2016,vats:fleg:jon:2019}. However, there is no unified approach to constructing estimators of $\Gamma(k)$ for parallel-chain implementations. Specifically, parallel chains are often spread across the state space so as to adequately capture high density areas. For slow mixing chains, the chains take time to traverse the space so that $\bar{X}_s$  are all vastly different. Consequently, $\hat{\Gamma}_s(k)$ typically underestimates $\Gamma(k)$ leading to a false sense of security about the quality of process.

@@

Estimators of ACvFs are used to estimate the long-run variance of Monte Carlo averages. Specifically, spectral variance (SV) estimators  are used to estimate $\Sigma = \sum_{k=-\infty}^{\infty} \Gamma(k)$ \citep{andr:1991,dame:1991,fleg:jone:2010}. We replace $\hat{\Gamma}_s(k)$ with G-ACvF in SV estimators to obtain globally-centered SV (G-SV) estimator and demonstrate strong consistency under weak conditions. In the spirit of \cite{andr:1991}, we also obtain large-sample bias and variance of the resulting estimator. SV estimators can be prohibitively slow to  compute \citep{liu:fleg:2018}. To relieve the computational burden, we adapt the fast algorithm of \cite{heberle2017fast} for G-SV estimator that dramatically reduces computation time. The G-SV estimator  is employed in the computation of ESS. We will show that using G-SV estimator for estimating $\Sigma$ safeguards users against early termination of the MCMC process.

@@

In Figure~\ref{fig:gaussian-acf} for $n = 10^4$, we present the ACF plots using $\hat{\Gamma}_s(k)$ and our proposed G-ACF estimator. The blue curves are the respective estimates at $n = 10^5$. At $n = 10^5$ when the chains have similar means, the G-ACF and locally-centered ACF are equivalent.  However, for $n = 10^4$, $\hat{\Gamma}_s(k)$ critically underestimate the correlation giving user a misleading visual of the quality of the Markov chain. 
The G-ACF estimator accounts for the discrepancy in sample means between two chains leading to a far improved quality of estimation.

\begin{figure}[htbp]
\centering
   \includegraphics[width=.35\linewidth]{plots/gaussian-Targettrace_n1e4.pdf}
    \includegraphics[width=.35\linewidth]{plots/gaussian-Targettrace_n1e5.pdf} 
    \caption{Target density and trace plots for two chains at $n = 10^4$ (left) and $n = 10^5$ (right)}
    \label{fig:gaussian-trace}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width=.6\textwidth]{plots/gaussian-acf_hist.pdf} \\
% \includegraphics[width=.6\textwidth]{plots/gaussian-acf_1e4.pdf}

\caption{ACF plots using local-centering (left) and global-centering (right). Histogram estimates at $n = 10^4$ with the blue curve being estimates from $n = 10^5$ for the first chain.}
 % (b) $n = 10^4$; the dotted lines are ACF for each chain and the thick solid line is the ACF averaged over both chains.}
\label{fig:gaussian-acf}
\end{figure}

@@

Let $P$ denote the Markov transition kernel that uniquely determines the ACvF under stationarity. Recall that the sample mean of the $s^{th}$ Markov chain is $\bar{X}_s$ and denote the global mean vector by $\bar{\bar{X}} = m^{-1}\sum_{s = 1}^{m}\bar{X}_s$. The global mean is a naturally superior estimator of $\mu$ compared to $\bar{X}_s$. For $k \geq 0$, we define the globally-centered ACvF (G-ACvF) estimator for the $s${th} Markov chain as
%
\begin{equation} \label{eq:G-ACvF}
    \hat{\Gamma}_{G,s}(k) = \dfrac{1}{n} \sum_{t=1}^{n-k}(X_{st}-\bar{\bar{X}})(X_{s(t+k)}-\bar{\bar{X}})^T,
\end{equation}
with $\hat{\Gamma}_{G,s}(k) = \hat{\Gamma}_{G,s}(-k)^T$ for $k < 0$. In the event that all $m$ Markov chains have been run long enough, $\bar{X}_s \approx \bar{\bar{X}}$, and hence  $\hat{\Gamma}_{s} (k) \approx \hat{\Gamma}_{G,s}(k)$. However, for shorter runs or for slow mixing chains, $\Gamma(k)$ is more appropriately estimated by $\hat{\Gamma}_{G,s}$ as it utilizes information from all chains and accounts for disparity between estimates of $\mu$. 
% We quantify the gains in  bias  for the G-ACvF estimator below. 
Let 
\begin{equation*}
\Phi^{(q)} = \sum_{ k= -\infty}^{\infty}\abs{k}^q \Gamma(k)\,,  
\end{equation*}
%
and let $\Phi^{(1)}$ be denoted by $\Phi$.  The proof of the following theorem is available in Appendix~\ref{appendix:bias}. Let $\|\cdot\|$ denote Euclidean norm.

@@

Let $\Gamma^{ii}(k)$ be the $i${th} component of $\Gamma(k)$. For component $i$, the autocorrelation is defined as
\[
\rho^{(i)}(k) = \dfrac{\Gamma^{ii}(k)}{\Gamma^{ii}(0)}\,,
\]
and is instrumental in visualizing the serial correlation in components of the Markov chain. A standard estimator of the autocorrelation is constructed from $\hat{\Gamma}_s(k)$. Instead, we advocate for using G-ACvF estimates so that,
\begin{equation}
\label{eq:acf}
 \hat{\rho}_{G,s}^{(i)}(k) = \dfrac{ \hat{\Gamma}^{(i)}_{G,s} (k)}{\hat{\Gamma}^{(i)}_{G,s} (0)}. 
\end{equation}
The globally-centered autocorrelation provides a far more realistic assessment of the correlation structure of marginal components of the chain  as evidenced in Figure~\ref{fig:gaussian-acf}.

>>>>

\onehalfspacing
\begin{abstract}
Autocovariances are a fundamental quantity of interest in Markov chain Monte Carlo (MCMC) simulations with autocorrelation function (ACF) plots being an integral performance assessment visualization tool. Unfortunately, for slow mixing Markov chains, the empirical autocovariance can highly underestimate the truth. For multiple chain MCMC sampling,  we propose a globally-centered estimator of the autocovariance function (G-ACvF) that exhibits significant theoretical and empirical improvements. We show that the bias of the G-ACvF estimator is smaller than the bias of the current state-of-the-art.  The impact of this improved estimator is evident in three critical output analysis applications: (1) ACF plots, (2) estimates of the Monte Carlo asymptotic covariance matrix, and (3) estimates of the effective sample size. Under weak conditions, we establish strong consistency of an improved asymptotic covariance estimator, and obtain its large sample bias and variance.  The performance of the new G-ACvF estimator is demonstrated through various examples.
\end{abstract}

@@

Let $F$ be the target distribution with mean $\mu$ defined on a space $\X \subseteq \mathbb{R}^d$ equipped with a countably generated $\sigma$-field $\mathcal{B}(\X)$. For $s = 1, \dots, m$, let $\{X_{st}\}_{t\geq1}$ be the $s${th} Harris ergodic $F$-invariant Markov chain \citep[see][for definitions]{meyn:twee:2009} employed to learn characteristics about $F$. The process is covariance stationary so that
the ACvF at lag $k$ is defined as 
%
\[
    \Gamma(k) = \Cov_F(X_{s1}, X_{s\,1+k})= \mathbb{E}_F \left[(X_{s1} - \mu)(X_{s\,1+k} - \mu)^T \right]\,.
\]
Estimating $\Gamma(k)$ is critical to assessing the quality of the sampler and the reliability of Monte Carlo estimators. Let $\bar{X}_s = n^{-1} \sum_{t=1}^{n} X_{st}$ denote the Monte Carlo estimator of $\mu$ from the $s$th chain. The standard estimator for $\Gamma(k)$ is the sample autocovariance matrix at lag $k \geq 0$:
%
\begin{equation} \label{eq:empirical_ACvF}
    \hat{\Gamma}_s(k) = \dfrac{1}{n}\sum_{t=1}^{n-k} \left(X_{st} - \bar{X}_s \right) \left(X_{s\, t + k} - \bar{X}_s \right)^T\,,
\end{equation}
and for $k < 0$, $\hat{\Gamma}_s(k) = \hat{\Gamma}_s(-k)^T$. For a single-chain MCMC run, the estimator $\hat{\Gamma}_s(k)$ is used to construct ACF plots, to estimate the long-run variance of Monte Carlo estimators \citep{hannan:1970,dame:1991}, and to estimate effective sample size (ESS) \citep{kass:carlin:gelman:neal:1998,gong:fleg:2016,vats:fleg:jon:2019}. However, there is no unified approach to constructing estimators of $\Gamma(k)$ for parallel-chain implementations. Specifically, parallel chains are often spread across the state space so as to adequately capture high density areas. For slow mixing chains, the chains take time to traverse the space so that $\bar{X}_s$  are all vastly different. Consequently, $\hat{\Gamma}_s(k)$ typically underestimates $\Gamma(k)$ leading to a false sense of security about the quality of the process.

@@

Estimators of ACvFs are used to estimate the long-run variance of Monte Carlo averages. Specifically, spectral variance (SV) estimators  are used to estimate $\Sigma = \sum_{k=-\infty}^{\infty} \Gamma(k)$ \citep{andr:1991,dame:1991,fleg:jone:2010}. We replace $\hat{\Gamma}_s(k)$ with G-ACvF in SV estimators to obtain a globally-centered SV (G-SV) estimator and demonstrate strong consistency under weak conditions. In the spirit of \cite{andr:1991}, we also obtain large-sample bias and variance of the resulting estimator. SV estimators can be prohibitively slow to  compute \citep{liu:fleg:2018}. To relieve the computational burden, we adapt the fast algorithm of \cite{heberle2017fast} for G-SV estimator that dramatically reduces computation time. The G-SV estimator  is employed in the computation of ESS. We will show that using G-SV estimator for estimating $\Sigma$ safeguards users against early termination of the MCMC process.

@@

In Figure~\ref{fig:gaussian-trace} for $n = 10^4$, we present the ACF plots using $\hat{\Gamma}_s(k)$ and our proposed G-ACvF estimator. The blue curves are the respective estimates at $n = 10^5$. At $n = 10^5$ when the chains have similar means, the G-ACF and locally-centered ACF are equivalent.  However, for $n = 10^4$, $\hat{\Gamma}_s(k)$ critically underestimates the correlation giving users a misleading visual of the quality of the Markov chain. 
The G-ACF estimator accounts for the discrepancy in sample means between two chains leading to a far improved quality of estimation.

\begin{figure}[htbp]
\centering
   \includegraphics[width=.62\linewidth]{plots/gaussian-Targettrace.pdf}\\
\vspace{-.5cm}
    \includegraphics[width=.62\textwidth]{plots/gaussian-acf_hist.pdf} 
    \caption{Top: Target density and trace plots for two chains at $n = 10^4$ (left) and $n = 10^5$ (right). Bottom: ACF plots for the first chain using local-centering (left) and global-centering (right). Histogram estimates at $n = 10^4$ with the blue curve being estimates from $n = 10^5$.}
    \label{fig:gaussian-trace}
\end{figure}

% \begin{figure}[htbp]
% \centering
% \includegraphics[width=.6\textwidth]{plots/gaussian-acf_hist.pdf} \\
% % \includegraphics[width=.6\textwidth]{plots/gaussian-acf_1e4.pdf}

% \caption{ACF plots using local-centering (left) and global-centering (right). Histogram estimates at $n = 10^4$ with the blue curve being estimates from $n = 10^5$ for the first chain.}
 % (b) $n = 10^4$; the dotted lines are ACF for each chain and the thick solid line is the ACF averaged over both chains.}
% \label{fig:gaussian-acf}
% \end{figure}

@@

Let $P$ denote the Markov transition kernel that uniquely determines the ACvF under stationarity. Recall that the sample mean of the $s${th} Markov chain is $\bar{X}_s$ and denote the global mean vector by $\bar{\bar{X}} = m^{-1}\sum_{s = 1}^{m}\bar{X}_s$. The global mean is a naturally superior estimator of $\mu$ compared to $\bar{X}_s$. For $k \geq 0$, we define the globally-centered ACvF (G-ACvF) estimator for the $s${th} Markov chain as
%
\begin{equation} \label{eq:G-ACvF}
    \hat{\Gamma}_{G,s}(k) = \dfrac{1}{n} \sum_{t=1}^{n-k}(X_{st}-\bar{\bar{X}})(X_{s(t+k)}-\bar{\bar{X}})^T,
\end{equation}
with $\hat{\Gamma}_{G,s}(k) = \hat{\Gamma}_{G,s}(-k)^T$ for $k < 0$. In the event that all $m$ Markov chains have been run long enough, $\bar{X}_s \approx \bar{\bar{X}}$, and hence  $\hat{\Gamma}_{s} (k) \approx \hat{\Gamma}_{G,s}(k)$. However, for shorter runs or for slow mixing chains, $\Gamma(k)$ is more appropriately estimated by $\hat{\Gamma}_{G,s}$ as it utilizes information from all chains and accounts for disparity between estimates of $\mu$. 
% We quantify the gains in  bias  for the G-ACvF estimator below. 
Let 
\begin{equation*}
\Phi^{(q)} = \sum_{ k= -\infty}^{\infty}\abs{k}^q \Gamma(k)\,,  
\end{equation*}
%
and let $\Phi^{(1)}$ be denoted by $\Phi$.  The proof of the following theorem is in Appendix~\ref{appendix:bias}. Let $\|\cdot\|$ denote Euclidean norm.

@@

Let $\Gamma^{ii}(k)$ and $\hat{\Gamma}_{G,s}^{ii}(k)$ be the $i${th} component of $\Gamma(k)$ and $\hat{\Gamma}_{G,s}(k)$, respectively. For component $i$, the autocorrelation is defined as
\[
\rho^{(i)}(k) = \dfrac{\Gamma^{ii}(k)}{\Gamma^{ii}(0)}\,,
\]
and is instrumental in visualizing the serial correlation in the components of the Markov chain. A standard estimator of the autocorrelation is constructed from $\hat{\Gamma}_s(k)$. Instead, we advocate for using G-ACvF estimates so that,
\begin{equation}
\label{eq:acf}
 \hat{\rho}_{G,s}^{(i)}(k) = \dfrac{ \hat{\Gamma}^{ii}_{G,s} (k)}{\hat{\Gamma}^{ii}_{G,s} (0)}. 
\end{equation}
The globally-centered autocorrelation provides a far more realistic assessment of the correlation structure of the marginal components of the chain  as evidenced in Figure~\ref{fig:gaussian-trace}.
